### 2018-06-13 Writing/response_to_reviewers.tex

\item{\textcolor{blue}{1. To be a metric, there are several properties that need to be satisfied. The proposed function
is positively valued and symmetric (copied below from line 178): [...] But, the following metric property does not hold for all x; y:}}
\label{proof}

We are really grateful that this reviewer picked up this flow in our mathematical demonstration of CD metric.
This problem is due to a lack of indications from our side and we apologise for this.

\subsubsection{Character states translation}
In fact, before being plugged into the CD equation, characters needs to be modified following Felsenstein's \textit{xyz} notation.
In other words, because the character states tokens are not representing any meaningful numerical value if the characters are unordered (e.g. there is no assumption that state 0 should come before state 1), they need to be reordered in a standard way so that the first token in the character is renamed to ``1'', the second to ``2'' etc.
For example, in a 3 state character for four taxa: \texttt{X = {1, 0, 2, 1}}, the character tokens that are integers representing the states could equally have been \texttt{X = {E, A, I, E}} or \texttt{X = {blue, red, absent, blue}}.
Therefore, to compare characters fairly, the tokens needs to be translated in a standard way such as the first token is replaced by the symbol ``1'', etc.
In this example \texttt{X = {1, 0, 2, 1}} becomes \texttt{X' = {1, 2, 3, 1}}
In the reviewer's demonstration, \texttt{x  = {1}} and \texttt{y  = {0}} thus become \texttt{x' = {1}} and \texttt{y' = {1}}.
Hence $CD_{(x',y')} = 0 \Leftrightarrow x' = y'$.

\subsubsection{CD as a metric to compare the splits between characters}
Furthermore, the CD metric, used in this manuscript as a proxy for character correlation actually measures the difference between in characters in terms of phylogenetic signal.
The character \texttt{X} described above implies 3 splits for four taxa (and of course, the character \texttt{X'} implies the exact same splits).
Another character \texttt{Y = {1, 1, 1, 2}} will imply only two splits, non shared with \texttt{X} and a character \texttt{Z = {0, 2, 1, 0}} (\texttt{Z' = {1, 2, 3, 1}}) would imply the same splits as \texttt{X} (hence, $CD_{(X',Z')} = 0$ and $CD_{(X',Y')} = 0.5$).
In other words the probability of X and Z generating conflicting split signal is null and this probability is of 0.5 for X and Y (i.e. half of the splits will be conflicting).

@@

Following these two points, the reviewer counter example $CD_{(x,y)} = 0$ means that 1) $x'$ and $y'$ are identical and the characters $x$ and $y$ imply the same number of splits in a tree (here no splits since only one taxon is used - however, this would hold for any $n$ number of taxa).
We have clarified these two points in the manuscript and in the mathematical demonstration.

We've added more explanations in the supplementary materials 1 on (1) the character translation, (2) the four-point metric proof and (3) on how to interpret the $CD$ metric.
Additionally, we added more precisions in the text:

@@

\textit{(i.e. entirely correlated characters would define identical splits between a set of taxa)}. lines @@@

\textit{Similarly, an hypothetical comparison of two character for only one taxa \texttt{x = \{0\}} and \texttt{y = \{1\}} would result in comparing two identical characters \texttt{x' = \{1\}} and \texttt{y' = \{1\}}.} lines @@@

\item{\textcolor{blue}{While some details of the simulation study are included, some important ones are missing. In particular, in the design of the study, how did you avoid or mitigate errors introduced by the programs simulating data? All of the programs use random-number generators and have their own biases. How have you accounted for this in the study of design to conclude that your results are due to signal, and not bias from the programs used?}}

%TG: really not sure about this comment... This is an epistemological point I think and I don't think we can solve this problem, even using true random-number generator. The error due to chaos is by definition impossible to take into account...

We do agree that generating true randomness in current software is tricky but we are not sure how to answer this comment.
In our simulations, random numbers where used at different levels:

\begin{enumerate}
    \item The generation of the ``true'' tree.
    \item The generation of the discrete morphological characters (each 100, 350 and 1000).
    \item The removal of the characters (for the ``randomised'' matrices).
    \item The starting seeds for the tree inferences (in both MrBayes and PAUP*).
\end{enumerate}

Because each of this steps are dependent in a cascade (i.e. the random seed used to generated the ``true'' tree influence the generation of the discrete characters which influences the ``randomised'' matrix which influences the results from the random starting seed in the tree inference).
However, we consider the influence of these cascading random numbers negligible in this analysis.
Furthermore, we assumed that the multiplication of negligible effects will result in a near null error due to the pseudo random-number generators.
In fine, our results are bound to be somehow biased by our prior believes and experience as well as by the tools (hardware and software) and the methods used in this manuscript but this is more of an epistemological point and is not specific to this study.

### 2018-06-13 Writing/response_to_reviewers.tex

\begin{enumerate}
    \item The generation of the ``true'' tree.
    \item The generation of the discrete morphological characters (each 100, 350 and 1000).
    \item The removal of the characters (for the ``randomised'' matrices).
    \item The starting seeds for the tree inferences (in both MrBayes and PAUP*).
\end{enumerate}

Because each of this steps are dependent in a cascade (i.e. the random seed used to generated the ``true'' tree influence the generation of the discrete characters which influences the ``randomised'' matrix which influences the results from the random starting seed in the tree inference).
However, we consider the influence of these cascading random numbers negligible in this analysis.
Furthermore, we assumed that the multiplication of negligible effects will result in a near null error due to the pseudo random-number generators.
In fine, our results are bound to be somehow biased by our prior believes and experience as well as by the tools (hardware and software) and the methods used in this manuscript but this is more of an epistemological point and is not specific to this study.

\item{\textcolor{blue}{How did you ``randomly'' modify your matrices? Did you sample under some distribution? If so, which one?}}

Every replaced characters where randomly selected from a uniform distribution (0,1). We have now specified this in the text:

\textit{Each replaced characters (i.e. the characters with a weight of +$1$) where randomly selected from the remaining characters using a uniform distribution (0,1).} lines @@@

@@

\item{\textcolor{blue}{It is not obvious why generated character sequences are representative of biological sequences or sample the space overall. Were the parameters chosen for the HKY model to fit a particular evolutionary scenario? Are they meant to capture all of the space?}}

There is actually a vigorous debate surrounding how to generate biologically meaningful morphological characters \citep[][;confidential reviews and personal communications witht he authors]{OReilly20160081,goloboff2017weighted,puttick2017uncertain,GoloboffEmpirical,OReillyEmpirical}.
This ongoing debate, however, focuses mainly on inference methods where data simulation is a tangential aspect of it.
Therefore, we decided to base our simulation protocol on the empirical observations showing that half of 206 analysed datasets supported a M$k$-like model \citep[][Fig. 3]{Wright01072016} and the other half being a combination of models not explicitly described here but that are believed to be covered with the HKY-binarised model \citep{puttick2017uncertain,OReillyEmpirical}.
In our simulations, the idea was to cover a broad range of morphological characters (with variable state numbers, evolution and transition rates) in order to capture a realistic image of the ``discrete morphological character space''.

%TG: add some text to the manuscript parphrasing this? It's kind of already written in there...

@@

\item{\textcolor{blue}{Similarly, why did you choose 0.85 for the Mk model? Does it represent some important set of character sequences? How does it represent the space in general?}}

This 0.85 proportion of binary characters and 0.15 of three state ones was extracted from the 100 empirical matrices analysed in \cite{Guillerme2016146} (Supplementary appendix 1. Appendix A: Tree inference) and \cite{ZouConvergence}.
We've specified this is the text:

\textit{We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters (from empirical observations in: Guillerme and Cooper, 2016b; Zou and Zhang, 2016).}

\item{\textcolor{blue}{There are statements that seem contradictory, like (line 335): ``Thus, additionally to the Wilcoxon test results, we considered distribution to be significantly similar if they had an overlap probability $>$ 0.95 and different if they had an overlap probability $>$ 0.05.''}}

### 2018-06-13 Writing/draft_highlighted.tex
advanced reviewer 2 comments

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The \hl{``morphological HKY-binary'' model} \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use an M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    However, \cite{Wright01072016} \hl{have shown that equal transition rate is a reasonable assumption for many empirical datasets.}
\end{itemize}

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (i.e. with an average character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (i.e. with an average character difference $<0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification (i.e. randomly weighting characters).
Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of 1260 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

>>>>

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The \hl{``morphological HKY-binary'' model} \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use an M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep[\hl{from empirical observations in: }][]{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    However, \cite{Wright01072016} \hl{have shown that equal transition rate is a reasonable assumption for many empirical datasets.}
\end{itemize}

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (i.e. with an average character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (i.e. with an average character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification (i.e. randomly weighting characters).
\hl{Each replaced characters (i.e. the characters with a weight of +$1$) where randomly selected from the remaining characters using a uniform distribution (0,1).}
Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of 1260 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})
